An Empirical Study of Invasion and Resistance for Iterated Prisoner's Dilemma Strategies

### 2017-05-11 main.tex
Adjust tex file to match new results.

All of our training code is available in the
Axelrod repository with documentation to train new strategies easily. Training
% TODO Include link etc
typically takes less than 200 iterations and can be completed within several
hours on commodity hardware.  The three strategies trained specifically for
this study are Trained FSM 1, 2, and 3 (TF1 - TF3), based on finite state
machines of 16, 16, and 8 states respectively (see Figures \ref{}).
% TODO Include images of FSMs
These strategies were trained with the objective function of mean fixation
probabilities for Moran processes starting at initial population states
consisting of \(N/2\) individuals of the training candidates and \(N/2\)
individuals of an opponent strategy, for all of the short run time strategies
in the library (approximately 150 opponents at the time of training):

\begin{itemize}
	\item TF1 \(N=12\), 0\% noise, 10000 repetitions per match
	\item TF2 \(N=10\), 0\% noise, 10000 repetitions per match
	\item TF3 \(N=8\), 1\% noise, 100 repetitions per match
\end{itemize}

@@

TF1 has an initial handshake of CCD and cooperates if the opponent matches.
However if the opponent later defects, TF1 will respond in kind, so the
handshake is not permanent. Only one player (Prober 4 \cite{prison}) manages to
achieve cooperation with TF1 after about 20 rounds of play.

@@

TF3 cooperates and defects with
various cycles depending on the opponent's actions. TF3 will mutually
cooperate with any strategy and only tolerates a few defections before
defecting for the rest of match. It is similar to but not exactly the same as
Fool Me Once, a strategy that cooperates until the opponent has defected twice
(not necessarily consecutively), and defects indefinitely thereafter. Though a
product of training with a Moran objective, it was selected for inclusion
because of the lack of an apparent handshake.

For both TF1 and TF2 a handshake
mechanism naturally emerges from the structure of the underlying finite state
machine. This behavior is an outcome of the evolutionary process and is in no
way hard-coded or included via an additional mechanism. Some of the existing
strategies in the library were trained on earlier versions of the library
(smaller total opponents used in training) based on neural networks, lookup
tables and stochastic variants trained with a particle swarm algorithm, finite
state machines and a stochastic variant mimicking hidden Markov models, and
neural networks. These strategies were trained to win IPD tournaments using an
objective function that computes the mean match payoffs against a collection of
opponents. Detailed descriptions and the performance of these strategies in IPD
tournaments will be described in another manuscript.
% TODO Think of how to reference this upcoming manuscript
% TODO Move the section below to the discussion for particular
% cases and/or conclusion
These strategies trained with the mean score
objective are among the best invaders in the library but are not as resistant
to invasion as the strategies trained using a Moran objective function (see
Tables \ref{}). These strategies include trained finite state machine
strategies, but they do not appear to have handshaking mechanisms. Therefore it
is reasonable to conclude that the objective function is the cause of the
emergence of handshaking mechanisms.
The payoff maximizing strategies
typically will not defect before the opponent's first defection, possibly
because the training strategy collection contains a significant portion of
strategies such as Grudger and Fool Me Once that retaliate harshly by defecting
for the remainder of the match if the opponent has more than a small number of
cumulative defections. Paradoxically it is necessary to defect in order to
achieve mutual cooperation with opponents using the same strategy but not with
other opponents. A handshake requires at least one defection and there is
selective pressure to defect as few times as possible to achieve the
self-recognition mechanism. It is also unwise to defect on the first move as
some strategies additionally retaliate first round defections. So the
handshakes used by TF1 and TF2, and CS, are in some sense optimal.  These
discoveries may have significant ramifications regarding the evolution of
cooperation and forgiveness in biological organisms and social interactions
between humans.

@@

Each strategy pair is run for 1000 runs of the Moran process to fixation
with starting population distributions of $(1, N-1)$,
$(N/2, N/2)$ and $(N-1 , 1)$, for \(N\) from 2 through 14. The
% TODO Change 14 to 12 if we do not use the 14 case.
fixation probability is then empirically computed for each combination of
starting distribution and value of \(N\).

@@

Figure~\ref{fig:number_of_stochastic_match_outcomes} shows the distribution of
the number of outcomes between all strategy pairs.
% TODO is this true? the distribution also matters, not just the number of outcomes
% TODO Good point, we could (if we wanted to) go down the line of checking
% specific distributions etc or perhaps we just remove and point at the good
% fits in the Validation section as a justification.

@@

\begin{itemize}
    \item Section~\ref{sec:two_individuals} considers the specific case of
        \(N=2\).
    \item Section~\ref{sec:strong_invaders} investigates the effect of
        population size on the ability of a strategy to invade another
        population. This will highlight how complex strategies with long
        memories outperform simpler strategies.
    \item Section~\ref{sec:strong_resistors} similarly investigates the
        ability to defend against an invasion.
    \item Section~\ref{sec:population_size} investigates the relationship
        between performance for differing population sizes.
    \item Section~\ref{sec:relative_fitness} calculates the relative fitness of all
        strategies.
\end{itemize}

@@

\begin{enumerate}
    \item The top strategy is the Collective Strategy (CS) which has a simple
        handshake mechanism (a cooperation followed by a defection on the first
        move). As long as the opponent plays the same handshake and does not
        defect in the future it cooperates. Otherwise it defects for all
        rounds~\cite{Li2009}. This strategy was specifically designed for
        evolutionary processes.
    \item The finite state machine strategy % TODO Perhaps refer to earlier
        % discussion in the methodology section about these FSM strategies
    \item The Defector: it always defects. As it has little potential
        interaction with itself, recall that \(N=2\) is considered, its
        aggressiveness is rewarded.
    \item The Aggravater strategy which plays like Grudger (responding to any
        defections with unconditional defections throughout) however starts by
        playing 3 defections.
    \item Predator, a finite state machine described in~\cite{Ashlock2006}.
\end{enumerate}

@@

\begin{itemize}
    \item Prober 4, complex strategy with an initial 20 move sequence of
        cooperations and defections~\cite{prison}. This initial sequence serves
        as approximate handshake.
    \item Grudger (which only performs well for \(N=3\)), starts by cooperating
        but will defect if at any point the opponent has defected.
    \item MEM2, an infinite memory strategy that switches between TfT, Tf2T, and
        Defector~\cite{Li2014}.
    \item Fool Me Once, forgives one defection but defects forever once a second
        defection takes place.
    \item  PSO Gambler and Evolved Lookerup 2 2 2: are strategies that make use
        of a lookup table mapping the first 2 moves of the opponent as well as
        the last 2 moves of both players to an action. The PSO gambler is a
        stochastic version which maps those states to probabilities of
        cooperating. The lookerup was described in \cite{Knight2016}.
	\item The evolved ANN strategies are neural networks that map a number of
		attributes (first move, number of cooperations, last move etc\dots) to
		an action. Both of these have been trained using an evolutionary
		algorithm and the ANN 5 was trained to perform well in a noisy
		tournament.
	\item The Evolved FSM 16, is a 16 state finite machine strategy that has
		been trained using an evolutionary algorithm.
\end{itemize}

As well as noting that the memory length and complexity of these strategies are
much greater than one, it is interesting to note that none of them are akin to
memory one strategies. Only one is stochastic.

@@

There are 3 strategies that are only top performers in \(x_{N-1}\):

\begin{itemize}
    \item Handshake: a slightly less aggressive version of the Collective
        strategy \cite{robson1989}. As long as the initial sequence is played
        then it cooperates. Thus it will do well in a population consisting of
        many members of itself: just as the Collective strategy does. However it
        is not aggressive enough to invade other populations.
    \item Winner 21: a strategy that makes its decision deterministically based
        on 1 round of its own strategy and 2 of the opponent's strategy
        \cite{Mathieu2015}.
	\item Predator: a finite state machine described in~\cite{Ashlock2006}
\end{itemize}

Interestingly none of these strategies are stochastic: this is explained by
the need of strategies to have a steady hand when interacting with their own
kind. In essence: acting stochastically increase the chance of friendly fire.
However we note that it is possible to design a strategy with a ``stochastic
handshake'' \cite{Lee2015}.

@@

>>>>

All of our training code is available in the
Axelrod repository with documentation to train new strategies easily. Training
% TODO Include link etc
typically takes less than 200 iterations and can be completed within several
hours on commodity hardware.  The three strategies trained specifically for
this study are Trained FSM 1, 2, and 3 (TF1 - TF3), based on finite state
machines of 8, 16, and 16 states respectively (see Figures \ref{}).
% TODO Include images of FSMs
These strategies were trained with the objective function of mean fixation
probabilities for Moran processes starting at initial population states
consisting of \(N/2\) individuals of the training candidates and \(N/2\)
individuals of an opponent strategy, for all of the short run time strategies
in the library (approximately 150 opponents at the time of training):

\begin{itemize}
	\item TF1 \(N=8\), 1\% noise, 100 repetitions per match
	\item TF2 \(N=10\), 0\% noise, 10000 repetitions per match
	\item TF3 \(N=12\), 0\% noise, 10000 repetitions per match
\end{itemize}

@@

TF1 cooperates and defects with
various cycles depending on the opponent's actions. TF1 will mutually
cooperate with any strategy and only tolerates a few defections before
defecting for the rest of match. It is similar to but not exactly the same as
Fool Me Once, a strategy that cooperates until the opponent has defected twice
(not necessarily consecutively), and defects indefinitely thereafter. Though a
product of training with a Moran objective, it was selected for inclusion
because of the lack of an apparent handshake.

@@

TF3 has an initial handshake of CCD and cooperates if the opponent matches.
However if the opponent later defects, TF3 will respond in kind, so the
handshake is not permanent. Only one player (Prober 4 \cite{prison}) manages to
achieve cooperation with TF3 after about 20 rounds of play.

For both TF2 and TF3 a handshake
mechanism naturally emerges from the structure of the underlying finite state
machine. This behavior is an outcome of the evolutionary process and is in no
way hard-coded or included via an additional mechanism. Some of the existing
strategies in the library were trained on earlier versions of the library
(smaller total opponents used in training) based on neural networks, lookup
tables and stochastic variants trained with a particle swarm algorithm, finite
state machines and a stochastic variant mimicking hidden Markov models, and
neural networks. These strategies were trained to win IPD tournaments using an
objective function that computes the mean match payoffs against a collection of
opponents. Detailed descriptions and the performance of these strategies in IPD
tournaments will be described in another manuscript.
% TODO Think of how to reference this upcoming manuscript

@@

Each strategy pair is run for 1000 runs of the Moran process to fixation
with starting population distributions of $(1, N-1)$,
$(N/2, N/2)$ and $(N-1 , 1)$, for \(N\) from 2 through 14. The
fixation probability is then empirically computed for each combination of
starting distribution and value of \(N\).

@@

Figure~\ref{fig:number_of_stochastic_match_outcomes} shows the distribution of
the number of outcomes between all strategy pairs.

@@

\begin{itemize}
    \item Section~\ref{sec:two_individuals} considers the specific case of
        \(N=2\).
    \item Section~\ref{sec:strong_invaders} investigates the effect of
        population size on the ability of a strategy to invade another
        population. This will highlight how complex strategies with long
        memories outperform simpler strategies.
    \item Section~\ref{sec:strong_resistors} similarly investigates the
        ability to defend against an invasion.
    \item Section~\ref{sec:population_size} investigates the relationship
        between performance for differing population sizes.
\end{itemize}

@@

\begin{enumerate}
    \item The top strategy is the Collective Strategy (CS) which has a simple
        handshake mechanism (a cooperation followed by a defection on the first
        move). As long as the opponent plays the same handshake and does not
        defect in the future it cooperates. Otherwise it defects for all
        rounds~\cite{Li2009}. This strategy was specifically designed for
        evolutionary processes.
    \item The Defector: it always defects. As it has little potential
        interaction with itself, recall that \(N=2\) is considered, its
        aggressiveness is rewarded.
    \item The Aggravater strategy which plays like Grudger (responding to any
        defections with unconditional defections throughout) however starts by
        playing 3 defections.
    \item Predator, a finite state machine described in~\cite{Ashlock2006}.
    \item Handshake: a slightly less aggressive version of the Collective
        strategy \cite{robson1989}. As long as the initial sequence is played
        then it cooperates. Thus it will do well in a population consisting of
        many members of itself: just as the Collective strategy does. However it
        is not aggressive enough to invade other populations.
\end{enumerate}

@@

\begin{itemize}
    \item Grudger (which only performs well for \(N=3\)), starts by cooperating
        but will defect if at any point the opponent has defected.
    \item MEM2, an infinite memory strategy that switches between TfT, Tf2T, and
        Defector~\cite{Li2014}.
    \item TF1, the finite state machine trained specifically for Moran processes
        described in Section~\cite{sec:introduction}.
    \item Prober 4, complex strategy with an initial 20 move sequence of
        cooperations and defections~\cite{prison}. This initial sequence serves
        as approximate handshake.
    \item  PSO Gambler and Evolved Lookerup 2 2 2: are strategies that make use
        of a lookup table mapping the first 2 moves of the opponent as well as
        the last 2 moves of both players to an action. The PSO gambler is a
        stochastic version which maps those states to probabilities of
        cooperating. The lookerup was described in \cite{Knight2016}.
	\item The evolved ANN strategies are neural networks that map a number of
		attributes (first move, number of cooperations, last move etc\dots) to
		an action. Both of these have been trained using an evolutionary
		algorithm and the ANN 5 was trained to perform well in a noisy
		tournament.
    \item The Evolved FSM 16 is a 16 state finite state machine trained to
        perform well in tournaments.
\end{itemize}

As well as noting that the memory length and complexity of these strategies are
much greater than one, it is interesting to note that none of them are akin to
memory one strategies. Only one is stochastic. The finite state machines trained
specifically for Moran processes do not appear in the top 5: whereas strategies
trained for tournament do. This is due to the nature of invasion: most of the
opponents will initially be different strategies.

The strategies trained with the mean score
objective are among the best invaders in the library but are not as resistant
to invasion as the strategies trained using a Moran objective function. These strategies include trained finite state machine
strategies, but they do not appear to have handshaking mechanisms. Therefore it
is reasonable to conclude that the objective function is the cause of the
emergence of handshaking mechanisms.
The payoff maximizing strategies
typically will not defect before the opponent's first defection, possibly
because the training strategy collection contains a significant portion of
strategies such as Grudger and Fool Me Once that retaliate harshly by defecting
for the remainder of the match if the opponent has more than a small number of
cumulative defections. Paradoxically it is necessary to defect in order to
achieve mutual cooperation with opponents using the same strategy but not with
other opponents. The importance of these handshakes will become apparent in the
next Section.

@@

There are are only two new strategies that appear in the top ranks for
\(x_{N-1}\): TF2 and TF3. These two strategies are with from CS the strongest
resistors. They all have handshakes, and whilst the handshake that CS has was
programmed, the handshakes of TF2 and TF3 evolved through an evolutionary
process. It is also worth noting that none of the highly trained strategies of
Section~\ref{sec:strong_invaders} perform as well as resistors. This indicates
the subtlety needed to recognise true opponents.

Interestingly none of these strategies are stochastic: this is explained by
the need of strategies to have a steady hand when interacting with their own
kind. In essence: acting stochastically increase the chance of friendly fire.
However it is possible to design a strategy with a ``stochastic
handshake'' \cite{Lee2015}.

A handshake requires at least one defection and there is
selective pressure to defect as few times as possible to achieve the
self-recognition mechanism. It is also unwise to defect on the first move as
some strategies additionally retaliate first round defections. So the
handshakes used by TF2 and TF3, and CS, are in some sense optimal.  These
discoveries may have significant ramifications regarding the evolution of
cooperation and forgiveness in biological organisms such as antibacterial
resistant bacteria and social interactions
between humans.

@@

% Discuss handshakes again. I think this is an important point.
